`r glue::glue('Doublet benchmarking results: {params$dataset}')`

### 2024-07-23 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
spacing to trigger styling

```{r}
plot_pca_calls <- function(df,
                           color_column,
                           plot_type = "calls",
                           title) {
  # Plot PCs colored by either:
  # if plot_type == "calls", color by singlet or doublet, showing doublets on top
  # if plot_type == "class", color by singlet, doublet, or ambiguous, showing doublets & ambiguous on top
  # df is expected to contain columns PC1, PC2, `color_column`, which should _not_ be provided as a string
  plot <- ggplot(df) +
    aes(
      x = PC1,
      y = PC2,
      color = {{color_column}}
    ) +
    geom_point(
      size = 0.75,
      alpha = 0.6
    ) +
    ggtitle(title) +
    guides(color = guide_legend(override.aes = list(size = 2)))

  if (plot_type == "calls") {
    plot <- plot +
      scale_color_manual(
        name = "",
        values = c("doublet" = "black", "singlet" = "gray90")) +
      geom_point(
        data = dplyr::filter(df, {{color_column}} == "doublet"),
        color = "black",
        size = 0.75
      )
  } else if (plot_type == "class") {
    plot <- plot +
      scale_color_manual(
        name = "Consensus",
        values = c("ambiguous" = "yellow",
                   "doublet" = "black",
                   "singlet" = "gray90")
      ) +
      geom_point(
        data = dplyr::filter(df, consensus_class != "singlet"),
        size = 0.75,
        alpha = 0.7,
        aes(color = consensus_class),
      )
  } else {
    stop("plot_type must be 'calls' or 'class'")
  }

@@

  ggplot(df) +
    aes(x = PC1,
        y = PC2,
        color = {{color_column}}) +
  geom_point(
    size = 0.75,
    alpha = 0.6
  ) +
  geom_point(
    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")),
    size = 0.75
  ) +
  scale_color_manual(name = "Call type", values = metric_colors) +
  ggtitle("Consensus call metrics") +
  guides(color = guide_legend(override.aes = list(size = 2)))
}
```

@@

```{r read_data}
# Read in data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {

      scdbl_tsv <- file.path(result_dir, glue::glue("{dataset}_scdblfinder.tsv"))
      scrub_tsv <- file.path(result_dir, glue::glue("{dataset}_scrublet.tsv"))
      sce_file <- file.path(data_dir, dataset, glue::glue("{dataset}.rds"))

@@

      scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
        tibble::rownames_to_column(var = "barcodes") |>
        dplyr::select(barcodes,
                      ground_truth = ground_truth_doublets,
                      PC1 = PCA.1,
                      PC2 = PCA.2) |>
        dplyr::left_join(
          scrub_df,
          by = "barcodes"
        ) |>
        dplyr::left_join(
          scdbl_df,
          by = "barcodes"
        )
    }
  ) |>
  purrr::set_names(dataset_names)
```

@@

```{r}
doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

      doublet_barcodes <- list(
        "scDblFinder" = df$barcodes[df$scdbl_prediction == "doublet"],
        "scrublet"    = df$barcodes[df$scrublet_prediction == "doublet"],
        "cxds"        = df$barcodes[df$cxds_prediction == "doublet"]
      )

      UpSetR::upset(fromList(doublet_barcodes), order.by = "freq") |> print()
      grid::grid.text( # plot title
        dataset,
        x = 0.65,
        y = 0.95,
        gp = grid::gpar(fontsize=16)
      )

    }
  )

```

@@

```{r}
doublet_df_list <- doublet_df_list |>
  purrr::map(
    \(dataset_df) {

      dataset_df |>
        dplyr::rowwise() |>

        dplyr::mutate(
          # Add column `consensus_call`
          # "doublet" if all three methods are doublet, and "singlet" otherwise
          consensus_call = dplyr::if_else(
            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
            "doublet",
            "singlet"
          ),
          # Add column `consensus_class`
          #  "doublet" if all three methods are doublet, "singlet" if all three methods are singlet,
          #  and "ambiguous" if there are any disagreements
          consensus_class = dplyr::case_when(
            consensus_call == "doublet" ~ "doublet",
            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "singlet") ~ "singlet",
            .default = "ambiguous"
          ),
          # Add column `confusion_call`
          #  tp/tn/fp/fn based on the `consensus_call` column
          confusion_call = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          )
        )
      }
  )
```

@@

```{r, fig.width = 12, fig.height = 9}
pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
                      "scrublet prediction" = "scrublet_prediction",
                      "cxds prediction" = "cxds_prediction",
                      "Ground truth" = "ground_truth",
                      "Consensus call" = "consensus_call")
doublet_df_list |>
  purrr::imap(
    \(df, dataset) {

      pc_plot_list <- pc_color_columns |>
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df,
              color_column = !!sym(color_column),
              plot_type = "calls",
              title = plot_title
            ) + theme(legend.position = "none")}
        )

@@

      patchwork::wrap_plots(pc_plot_list) +
        plot_annotation(
            glue::glue("PCA for {dataset}"),
            theme = theme(plot.title = element_text(size = 16))) +
        plot_layout(guides = "collect")
    }
  )
```

@@

        # run confusion matrix only if there are both singlets and doublets
        if (length(unique(df$consensus_call)) == 2) {
            confusion_result <- caret::confusionMatrix(
            # truth should be first
            table(
              "Truth" = df$ground_truth,
              "Consensus prediction" = df$consensus_call
            ),
            positive = "doublet"
          )

@@

```{r}
# modify existing consensus columns to only consider 2 methods
doublet_df_list <- doublet_df_list |>
  purrr::map(
    \(dataset_df) {

      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        # Update column `consensus_call`
        dplyr::mutate(
          consensus_call = dplyr::if_else(
            all(c(scdbl_prediction, cxds_prediction) == "doublet"),
            "doublet",
            "singlet"
          )
        ) |>
        # Add Update `consensus_class`
        dplyr::mutate(
          consensus_class = dplyr::case_when(
            consensus_call == "doublet" ~ "doublet",
            all(c(scdbl_prediction, cxds_prediction) == "singlet") ~ "singlet",
            .default = "ambiguous"
          )
         ) |>
        # Add Update `confusion_call`
        dplyr::mutate(
          confusion_call = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          )
        )

@@

```{r, fig.width = 12, fig.height = 6}
pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
                      "cxds prediction" = "cxds_prediction",
                      "Ground truth" = "ground_truth",
                      "Consensus call" = "consensus_call")

doublet_df_list |>
  purrr::imap(
    \(df, dataset) {

      pc_plot_list <- pc_color_columns |>
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df,
              color_column = !!sym(color_column),
              plot_type = "calls",
              title = plot_title
            ) + theme(legend.position = "none")}
        )
      # plot for consensus_class, which needs three colors
      # note that the list name isn't actually used here
      pc_plot_list$consensus_class <- plot_pca_calls(
        df,
        color_column = consensus_class,
        plot_type = "class",
        title = "Consensus class"
      )

>>>>

```{r}
plot_pca_calls <- function(df,
                           color_column,
                           plot_type = "calls",
                           title) {
  # Plot PCs colored by either:
  # if plot_type == "calls", color by singlet or doublet, showing doublets on top
  # if plot_type == "class", color by singlet, doublet, or ambiguous, showing doublets & ambiguous on top
  # df is expected to contain columns PC1, PC2, `color_column`, which should _not_ be provided as a string
  plot <- ggplot(df) +
    aes(
      x = PC1,
      y = PC2,
      color = {{ color_column }}
    ) +
    geom_point(
      size = 0.75,
      alpha = 0.6
    ) +
    ggtitle(title) +
    guides(color = guide_legend(override.aes = list(size = 2)))

  if (plot_type == "calls") {
    plot <- plot +
      scale_color_manual(
        name = "",
        values = c("doublet" = "black", "singlet" = "gray90")
      ) +
      geom_point(
        data = dplyr::filter(df, {{ color_column }} == "doublet"),
        color = "black",
        size = 0.75
      )
  } else if (plot_type == "class") {
    plot <- plot +
      scale_color_manual(
        name = "Consensus",
        values = c(
          "ambiguous" = "yellow",
          "doublet" = "black",
          "singlet" = "gray90"
        )
      ) +
      geom_point(
        data = dplyr::filter(df, consensus_class != "singlet"),
        size = 0.75,
        alpha = 0.7,
        aes(color = consensus_class),
      )
  } else {
    stop("plot_type must be 'calls' or 'class'")
  }

@@

  ggplot(df) +
    aes(
      x = PC1,
      y = PC2,
      color = {{ color_column }}
    ) +
    geom_point(
      size = 0.75,
      alpha = 0.6
    ) +
    geom_point(
      data = dplyr::filter(df, {{ color_column }} %in% c("fp", "fn")),
      size = 0.75
    ) +
    scale_color_manual(name = "Call type", values = metric_colors) +
    ggtitle("Consensus call metrics") +
    guides(color = guide_legend(override.aes = list(size = 2)))
}
```

@@

```{r read_data}
# Read in data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {
      scdbl_tsv <- file.path(result_dir, glue::glue("{dataset}_scdblfinder.tsv"))
      scrub_tsv <- file.path(result_dir, glue::glue("{dataset}_scrublet.tsv"))
      sce_file <- file.path(data_dir, dataset, glue::glue("{dataset}.rds"))

@@

      scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
        tibble::rownames_to_column(var = "barcodes") |>
        dplyr::select(barcodes,
          ground_truth = ground_truth_doublets,
          PC1 = PCA.1,
          PC2 = PCA.2
        ) |>
        dplyr::left_join(
          scrub_df,
          by = "barcodes"
        ) |>
        dplyr::left_join(
          scdbl_df,
          by = "barcodes"
        )
    }
  ) |>
  purrr::set_names(dataset_names)
```

@@

```{r}
doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {
      doublet_barcodes <- list(
        "scDblFinder" = df$barcodes[df$scdbl_prediction == "doublet"],
        "scrublet"    = df$barcodes[df$scrublet_prediction == "doublet"],
        "cxds"        = df$barcodes[df$cxds_prediction == "doublet"]
      )

      UpSetR::upset(fromList(doublet_barcodes), order.by = "freq") |> print()
      grid::grid.text( # plot title
        dataset,
        x = 0.65,
        y = 0.95,
        gp = grid::gpar(fontsize = 16)
      )
    }
  )
```

@@

```{r}
doublet_df_list <- doublet_df_list |>
  purrr::map(
    \(dataset_df) {
      dataset_df |>
        dplyr::rowwise() |>
        dplyr::mutate(
          # Add column `consensus_call`
          # "doublet" if all three methods are doublet, and "singlet" otherwise
          consensus_call = dplyr::if_else(
            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"),
            "doublet",
            "singlet"
          ),
          # Add column `consensus_class`
          #  "doublet" if all three methods are doublet, "singlet" if all three methods are singlet,
          #  and "ambiguous" if there are any disagreements
          consensus_class = dplyr::case_when(
            consensus_call == "doublet" ~ "doublet",
            all(c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "singlet") ~ "singlet",
            .default = "ambiguous"
          ),
          # Add column `confusion_call`
          #  tp/tn/fp/fn based on the `consensus_call` column
          confusion_call = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          )
        )
    }
  )
```

@@

```{r, fig.width = 12, fig.height = 9}
pc_color_columns <- c(
  "scDblFinder prediction" = "scdbl_prediction",
  "scrublet prediction" = "scrublet_prediction",
  "cxds prediction" = "cxds_prediction",
  "Ground truth" = "ground_truth",
  "Consensus call" = "consensus_call"
)
doublet_df_list |>
  purrr::imap(
    \(df, dataset) {
      pc_plot_list <- pc_color_columns |>
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df,
              color_column = !!sym(color_column),
              plot_type = "calls",
              title = plot_title
            ) + theme(legend.position = "none")
          }
        )

@@

      patchwork::wrap_plots(pc_plot_list) +
        plot_annotation(
          glue::glue("PCA for {dataset}"),
          theme = theme(plot.title = element_text(size = 16))
        ) +
        plot_layout(guides = "collect")
    }
  )
```

@@

      # run confusion matrix only if there are both singlets and doublets
      if ("singlet" %in% df$consensus_call & "doublet" %in% df$consensus_call) {
        confusion_result <- caret::confusionMatrix(
          # truth should be first
          table(
            "Truth" = df$ground_truth,
            "Consensus prediction" = df$consensus_call
          ),
          positive = "doublet"
        )

@@

```{r}
# modify existing consensus columns to only consider 2 methods
doublet_df_list <- doublet_df_list |>
  purrr::map(
    \(dataset_df) {
      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        # Update column `consensus_call`
        dplyr::mutate(
          consensus_call = dplyr::if_else(
            all(c(scdbl_prediction, cxds_prediction) == "doublet"),
            "doublet",
            "singlet"
          )
        ) |>
        # Add Update `consensus_class`
        dplyr::mutate(
          consensus_class = dplyr::case_when(
            consensus_call == "doublet" ~ "doublet",
            all(c(scdbl_prediction, cxds_prediction) == "singlet") ~ "singlet",
            .default = "ambiguous"
          )
        ) |>
        # Add Update `confusion_call`
        dplyr::mutate(
          confusion_call = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          )
        )

@@

```{r, fig.width = 12, fig.height = 6}
pc_color_columns <- c(
  "scDblFinder prediction" = "scdbl_prediction",
  "cxds prediction" = "cxds_prediction",
  "Ground truth" = "ground_truth",
  "Consensus call" = "consensus_call"
)

doublet_df_list |>
  purrr::imap(
    \(df, dataset) {
      pc_plot_list <- pc_color_columns |>
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df,
              color_column = !!sym(color_column),
              plot_type = "calls",
              title = plot_title
            ) + theme(legend.position = "none")
          }
        )
      # plot for consensus_class, which needs three colors
      # note that the list name isn't actually used here
      pc_plot_list$consensus_class <- plot_pca_calls(
        df,
        color_column = consensus_class,
        plot_type = "class",
        title = "Consensus class"
      )

### 2024-07-23 analyses/doublet-detection/template-notebooks/02_explore-benchmark-results.Rmd
spacing to trigger styling

# SCE, with ground truth included
sce_file <- file.path(
  data_dir,
  glue::glue("{params$dataset}.rds")
)

```

@@

  # If there are no predicted doublets, make sure the tab has a 0 count
  dub_tab <- table( factor(df[[pred_col]], levels = c("singlet", "doublet") ))
  dub_percent <- round(dub_tab[["doublet"]]/sum(dub_tab), 4) * 100

@@

  p <- ggplot(df) +
    aes(x = PC1,
        y = PC2,
        color = {{color_column}}) +
  geom_point(alpha = 0.5) +
  labs(color = color_lab) +
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )

  # Set the palette, and ensure doublets are on top for visibility
  if (color_type == "d") {
    p <- p +
      scale_color_manual(values = c("black", "lightblue")) +
    geom_point(
      data = dplyr::filter(df, {{color_column}} == "doublet"),
      color = "black"
    )
  } else if (color_type == "c") {
    p <- p +
      scale_color_viridis_c(direction = -1) +
    geom_point(
      data = dplyr::filter(df, {{pred_column}} == "doublet"),
      aes(color = {{color_column}})
    )
  }

@@

plot_jitter_scores <- function(df, score_col, pred_col, method_name){
  # Plot jitter plot of score distributions and predictions, colored by ground truth
  # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(x = ground_truth,
        y = {{score_col}},
        color = {{pred_col}}) +
    geom_jitter(
      alpha = 0.2,
      width = 0.1
    ) +
    labs(
      x = "Ground truth",
      y = glue::glue("{method_name} score"),
      color = glue::glue("{method_name} prediction")
    ) +
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )
}

plot_density_scores <- function(df, score_col, method_name){
  # Plot density plot of score distributions, colored by ground truth
  # df is expected to contain columns `ground_truth` and `score_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(x = {{score_col}},
        fill = ground_truth) +
    geom_density(alpha = 0.7) +
    labs(
      x = glue::glue("{method_name} score"),
      fill = "Ground truth call"
    ) +
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )
}

@@

    if (!all(dim(confusion_table) == c(2,2))) {
      print("Metrics could not be calculated; insufficient counts.")
      return(NULL)
    } else {
      return(
        caret::confusionMatrix(
          confusion_table,
          positive = "doublet"
        )
      )
    }

}

```

@@

# grab ground truth and PCA coordinates
sce <- read_rds(sce_file)
sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
  tibble::rownames_to_column(var = "barcodes") |>
  dplyr::select(barcodes,
                ground_truth = ground_truth_doublets,
                PC1 = PCA.1,
                PC2 = PCA.2)

@@

```{r, eval = has_scdbl, fig.width = 8}

d_plot <- plot_pca(
  doublet_df,
  color_column = scdbl_prediction,
  color_lab = "scDblFinder prediction"
)

@@

```{r fig.width = 10}

cxds_cols <- names(doublet_df)[grep("^cxds_prediction_", names(doublet_df))]

>>>>

# SCE, with ground truth included
sce_file <- file.path(
  data_dir,
  glue::glue("{params$dataset}.rds")
)
```

@@

  # If there are no predicted doublets, make sure the tab has a 0 count
  dub_tab <- table(factor(df[[pred_col]], levels = c("singlet", "doublet")))
  dub_percent <- round(dub_tab[["doublet"]] / sum(dub_tab), 4) * 100

@@

  p <- ggplot(df) +
    aes(
      x = PC1,
      y = PC2,
      color = {{ color_column }}
    ) +
    geom_point(alpha = 0.5) +
    labs(color = color_lab) +
    theme(
      legend.title.position = "top",
      legend.position = "bottom"
    )

  # Set the palette, and ensure doublets are on top for visibility
  if (color_type == "d") {
    p <- p +
      scale_color_manual(values = c("black", "lightblue")) +
      geom_point(
        data = dplyr::filter(df, {{ color_column }} == "doublet"),
        color = "black"
      )
  } else if (color_type == "c") {
    p <- p +
      scale_color_viridis_c(direction = -1) +
      geom_point(
        data = dplyr::filter(df, {{ pred_column }} == "doublet"),
        aes(color = {{ color_column }})
      )
  }

@@

plot_jitter_scores <- function(df, score_col, pred_col, method_name) {
  # Plot jitter plot of score distributions and predictions, colored by ground truth
  # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(
      x = ground_truth,
      y = {{ score_col }},
      color = {{ pred_col }}
    ) +
    geom_jitter(
      alpha = 0.2,
      width = 0.1
    ) +
    labs(
      x = "Ground truth",
      y = glue::glue("{method_name} score"),
      color = glue::glue("{method_name} prediction")
    ) +
    theme(
      legend.title.position = "top",
      legend.position = "bottom"
    )
}

plot_density_scores <- function(df, score_col, method_name) {
  # Plot density plot of score distributions, colored by ground truth
  # df is expected to contain columns `ground_truth` and `score_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(
      x = {{ score_col }},
      fill = ground_truth
    ) +
    geom_density(alpha = 0.7) +
    labs(
      x = glue::glue("{method_name} score"),
      fill = "Ground truth call"
    ) +
    theme(
      legend.title.position = "top",
      legend.position = "bottom"
    )
}

@@

  if (!all(dim(confusion_table) == c(2, 2))) {
    print("Metrics could not be calculated; insufficient counts.")
    return(NULL)
  } else {
    return(
      caret::confusionMatrix(
        confusion_table,
        positive = "doublet"
      )
    )
  }
}
```

@@

# grab ground truth and PCA coordinates
sce <- read_rds(sce_file)
sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
  tibble::rownames_to_column(var = "barcodes") |>
  dplyr::select(barcodes,
    ground_truth = ground_truth_doublets,
    PC1 = PCA.1,
    PC2 = PCA.2
  )

@@

```{r, eval = has_scdbl, fig.width = 8}
d_plot <- plot_pca(
  doublet_df,
  color_column = scdbl_prediction,
  color_lab = "scDblFinder prediction"
)

@@

```{r fig.width = 10}
cxds_cols <- names(doublet_df)[grep("^cxds_prediction_", names(doublet_df))]
